---
output:
  pdf_document: default
  html_document: default
  md_document:
    variant: markdown_github
editor_options:
  chunk_output_type: console
---

<!-- README.md is generated from CancerMortality.Rmd. Please edit that file. --> 

```{r opts, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "images/"
)
```

# Data Preparation
```{r, echo=FALSE}
library(crayon)
library(car)
library(FactoMineR)
library(chemometrics)
library(corrplot)
library(RColorBrewer)
library(PerformanceAnalytics)
library(mice)
library(dplyr)
library(readr)
```

First we import the data and save it as the variable "df" for future 
modifications. 

```{r}	
par(mfrow=c(1,1))
df <- read_csv("data/train.csv")
```

Here's a breakdown of all the variables in the dataset.

**Variable Name**       | **Num/Fac** | **Description**
 ---------------------- | ----------- | ---------------------- 
avganncount             | N           | Mean number of reported cases of cancer diagnosed annually (2010-2015 
avgdeathsperyear        | N           | Mean number of reported mortalities due to cancer 
target_deathrate        | N           | Response variable. Mean per capita (100,000) cancer mortalities 
incidencerate           | N           | Mean per capita (100,000) cancer diagnoses 
medincome               | N           | Median income per county 
popest2015              | N           | Population of county 
povertypercent          | N           | Percent of population in poverty 
studypercap             | N           | Per capita number of cancer-related clinical trials per county 
binnedinc               | F           | Median income per capita binned by decile 
medianage               | N           | Median age of county residents 
medianagemale           | N           | Median age of male county residents 
medianagefemale         | N           | Median age of female county residents 
geography               | F           | County name 
percentmarried          | N           | Percent of county residents who are married 
pctnohs18_24            | N           | Percent  of  county  residents  ages  18-24  highest  education  attained: less than high school  
pcths18_24              | N           | Percent  of  county  residents  ages  18-24  highest  education  attained:  high school diploma 
pctsomecol18_24         | N           | Percent  of  county  residents  ages  18-24  highest  education  attained: some college 
pcths25_over            | N           | Percent  of  county  residents  ages  25  and  over  highest  education attained: high school diploma 
pctbachdeg25_over       | N           | Percent  of  county  residents  ages  25  and  over  highest  education attained: bachelor's degree 
pctemployed16_over      | N           | Percent of county residents ages 16 and over employed 
pctunemployed16_over    | N           | Percent of county residents ages 16 and over unemployed 
pctprivatecoverage      | N           | Percent of county residents with private health coverage 
pctprivatecoveragealone | N           | Percent  of  county  residents  with  private  health  coverage  alone  (no public assistance) 
pctempprivcoverage      | N           | Percent  of  county  residents  with  employee-provided  private  health coverage 
pctpubliccoverage       | N           | Percent of county residents with government-provided health coverage 
pctpubliccoveragealone  | N           | Percent of county residents with government-provided health coverage alone 
pctwhite                | N           | Percent of county residents who identify as White 
pctblack                | N           | Percent of county residents who identify as Black 
pctasian                | N           | Percent of county residents who identify as Asian 
pctotherrace            | N           | Percent  of  county  residents  who  identify  in  a  category  which  is  not White, Black, or Asian 
pctmarriedhouseholds    | N           | Percent of married households 
birthrate               | N           | Number of live births relative to number of women in county


## Variable Types

All variables but `geography` should be numeric. Only one variable requires
any change: `binnedinc` is a string variable right now, but two things can be
created from it: a factor variable (`f.binnedinc`), and a numeric variable 
holding the midpoint of each bin, which we'll conviniently call `binnedinc`.

```{r}
df$f.binnedinc <- as.factor(df$binnedinc)
# Use regex to remove the [,],( and ) from the rows:
inc.midpoints.text <- gsub("[\\[\\]()]", "", df$binnedinc, perl = T)
# Separate them into two numbers
inc.midpoints.text.sep <- strsplit(inc.midpoints.text, ",")
# Convert them to numbers and apply a mean between them to find the midpoint
df$binnedinc <- sapply(inc.midpoints.text.sep, function(x) mean(as.numeric(x)))
```

Note how `geography`, altough being non-numeric, it's not a factor variable.
This can be shown by looking at the number of unique values in the column.

```{r}
nrow(df); length(unique(df$geography))
```

Because the number of unique values is the same as the number of rows, we can
safely assume that `geography` is a unique identifier for each row. It won't be
removed yet, because, as proven in a later section, a factor variable can be
derived from it.

## Duplicated Data Points

Before proceeding with the analysis, we should check if there are any duplicated
rows in the data set. If there are, we shall remove them.

```{r}
duplicated_row_count <- sum(duplicated(df))
if (duplicated_row_count > 0) {
    print(sprintf("There are %d duplicated rows.", duplicated_row_count))
    df <- unique(df)
}
```

No duplicated rows were found in the data set.

## Missing data

Glancing over the data set, one can see that there are some missing values:

```{r}
for (colname in colnames(df)) {
    na.count <- sum(is.na(df[[colname]]))
    if (na.count > 0) {
        cat(sprintf("%s has %s\n", colname, red(sprintf("%d N/As", sum(is.na(df[colname]))))))
    }
}
```

* Column `pctsomecol18_24` has 1376 N/As, which makes for more than 75% of the
  data. Due to that, it would most likely provide little to no meaningful
  information, so a decision was made to remove it from the study.
* Column `pctemployed16_over` has 82 N/As, which is a manageable amount. They
  can easily be imputed using the MICE method.
* Column `pctprivatecoveragealone` has 356 N/As, which acounts for less than 20%
  of the rows. This is a small enough amount to be imputed using the MICE 
  method. Nontheless, this column will be removed from the study later on, for
  reasons explained in the exploratory data analysis section.

```{r}
# Drop the column with too many missing values
df <- subset(df, select = -c(pctsomecol18_24))

# Impute missing values
res.mice <- mice(df)
complete_df = complete(res.mice, action = 1)
```

Before substituting the missing values, let's check the deciles of the variables
to see if the imputation makes sense.

```{r}
quantile(df$pctemployed16_over, na.rm = TRUE, probs = seq(0, 1, 0.1))
quantile(df$pctprivatecoveragealone, na.rm = TRUE, probs = seq(0, 1, 0.1))
```

Now let's substitute the missing values in the original data set. And check that
the deciles are still roughly the same. 

```{r}
df$pctemployed16_over <- complete_df$pctemployed16_over
df$pctprivatecoveragealone <- complete_df$pctprivatecoveragealone

quantile(df$pctemployed16_over, na.rm = TRUE, probs = seq(0, 1, 0.1))
quantile(df$pctprivatecoveragealone, na.rm = TRUE, probs = seq(0, 1, 0.1))
```

Which they are.

## Univariate Outliers

Column-wise, we can count how many univariate outliers each numeric variable 
has:

```{r}
for (colname in colnames(Filter(is.numeric, df))) {
  col = df[[colname]]
  q1 <- quantile(col, 0.25)
  q3 <- quantile(col, 0.75)
  iqr <- q3 - q1

  severe <- list(top = q3 + 3 * iqr, bot = q1 - 3 * iqr)
  mild <- list(top = q3 + 1.5 * iqr, bot = q1 - 1.5 * iqr)

  severe_out <- sum(col > severe$top | col < severe$bot)
  mild_out <- sum((col > mild$top & col < severe$top) | (col < mild$bot & col > severe$bot))
  if (mild_out > 0 | severe_out > 0) {
    cat(sprintf("Column %s has %d mild outliers and %d severe outliers\n", colname, mild_out, severe_out))
  }
}
```

Row-wise, we'll count the numeric variables in which each data point is an 
outlier, and create a new object called `univariate_outlier_count`. As a 
gut-driven criterion, we shall consider a row to be an outlier if it is an
outlier in 10 or more variables. Based on this criterion, only 9 counties are.

```{r}
count_outliers <- function(data) {
  # Function to check for outliers based on IQR
  is_outlier <- function(x) {
    Q1 <- quantile(x, 0.25, na.rm = TRUE)
    Q3 <- quantile(x, 0.75, na.rm = TRUE)
    IQR <- Q3 - Q1
    lower_bound <- Q1 - 1.5 * IQR
    upper_bound <- Q3 + 1.5 * IQR
    return(x < lower_bound | x > upper_bound)
  }
  
  # Apply the outlier function to each column and sum the results for each row using dplyr
  data %>%
    mutate(outlier_count = rowSums(sapply(., is_outlier), na.rm = TRUE))
}

univariate_outlier_count <- count_outliers(Filter(is.numeric, df))$outlier_count
df[which(univariate_outlier_count >= 10),]
```

All of them have high percentages of non-white population, both black and asian,
a low median age, a high mortality count and a high bias towards private and
employee health coverage. Of these 9 counties, 7 are wealthy (Low poverty
percent) and 2 have a large poor population (over 20%).

Outliers can sometimes provide valuable information, so they won't be removed
from the data set just yet.

## Multivariate Outliers

We will apply Moutlier on the numerical variables in order to find multivariate 
outliers. We have to perform the calculation excluding the variable studypercap 
because otherwise the method is unable to execute due to multicollinearity 
casuing a singularity matrix in the intermediate calculations. An extremely mild 
threshold is chosen (0.00005%) because even using this threshold we get a 
significant amount of multivariate outliers, 4% of the total sample. Lowering 
the threshold even further doesn't change much the amount of outliers and rising 
it higher makes the amount of outliers rise too much (10% outliers at 0.1% 
significance level).

```{r}
numeric.df <- Filter(is.numeric, df)
numeric.df <- numeric.df[, !colnames(numeric.df) %in% c("studypercap")]

res.out_95 <- Moutlier(numeric.df, quantile = 0.95, plot=F)
multi_outliers_95 = which((res.out_95$md > res.out_95$cutoff)&(res.out_95$rd > res.out_95$cutoff))
length(multi_outliers_95)

res.out <- Moutlier(numeric.df, quantile = 0.9999995, plot=F)
multi_outliers = which((res.out$md > res.out$cutoff)&(res.out$rd > res.out$cutoff))
length(multi_outliers)

par(mfrow = c(1,1))
plot(res.out$rd, res.out$md )
abline(h=res.out$cutoff, col="red")
abline(v=res.out$cutoff, col="red")
```

There are 91 multivariate outliers in the data set (265 if we take a 95% 
quantile).

# Exploratory Data Analysis

This section will be devided in two parts: single-variable analysis and
multi-variable analysis.

## Single-Variable analysis

This sub-section presents an analysis for each variable of the data set as a 
standalone sample.

We'll be performing lots of discretisation of continuous variables based on 
their quartiles, so let's create a function to do that.

```{r}
discretize_quartiles <- function(column, label_name) {
  res <- cut(column, breaks = quantile(column, probs = seq(0, 1, 0.25)), 
    include.lowest = T,
    labels=c(
      sprintf("Low%s", label_name),
      sprintf("LowMid%s", label_name),
      sprintf("HighMid%s", label_name),
      sprintf("High%s", label_name)
    )
  )
  print(table(res)) # Print the table
  return(res)
}
```

### Variable 1 - avganncount

This is a continuous ratio variable. The data does not look normally 
distributed, which is confirmed by the near-null p-value of the shapiro 
normallity test. A histogram is used to visualize the data. 

```{r}
summary(df$avganncount)

hist(df$avganncount, breaks = 30, freq = F)
curve(dnorm(x, mean(df$avganncount), sd(df$avganncount)), add = T)

shapiro.test(df$avganncount)
```

An additional factor `f.avganncount` is created to discretize the data according
to the quartiles.

```{r}
df$f.avganncount <- discretize_quartiles(df$avganncount, "CaseCount")
```

### Variable 2 - avgdeathsperyear

This is also a continuous ratio variable similar to variable 1. The data does 
not look normally distributed, which is confirmed by the near-null p-value of 
the shapiro normallity test. Again a histogram is used to visualize the data.


```{r}
summary(df$avgdeathsperyear)

hist(df$avgdeathsperyear, breaks = 30, freq = F)
curve(dnorm(x, mean(df$avgdeathsperyear), sd(df$avgdeathsperyear)), add = T)

shapiro.test(df$avgdeathsperyear)
```

An additional factor `f.avgdeathsperyear` is created to discretize the data
according to the quartiles.

```{r}
df$f.avgdeathsperyear <- discretize_quartiles(df$avgdeathsperyear, "MortCount")
```

### Variable 3 - target_deathrate

This is the response variable. This is also a continuous ratio variable similar 
to the previous variables. The data looks normally distributed, but it is not 
and will be further discussed in the next section. 

```{r}
summary(df$target_deathrate)

hist(df$target_deathrate, breaks = 30, freq = F)
curve(dnorm(x, mean(df$target_deathrate), sd(df$target_deathrate)), add = T)

shapiro.test(df$target_deathrate)
```

An additional factor `f.target_deathrate` is created to discretize the data
according to the quartiles.

```{r}
df$f.target_deathrate <- discretize_quartiles(df$target_deathrate, "DeathRate")
```

### Variable 4 - incidencerate

We have another continuous ratio variable similar to the previous variables. It 
is not normally distributed according to the Shapiro test.

```{r}
summary(df$incidencerate)

hist(df$incidencerate, breaks = 30, freq = F)
curve(dnorm(x, mean(df$incidencerate), sd(df$incidencerate)), add = T)

shapiro.test(df$incidencerate)
```

An additional factor `f.incidencerate` is created to discretize the data
according to the quartiles.

```{r}
df$f.incidencerate <- discretize_quartiles(df$incidencerate, "DiagnPerCap")
```

### Variable 5 - medincome

Very similar to all the previous variables we have a continuous ratio variable 
not normally distributed.

```{r}
summary(df$medincome)

hist(df$medincome, breaks = 30, freq = F)
curve(dnorm(x, mean(df$medincome), sd(df$medincome)), add = T)

shapiro.test(df$medincome)
```

An additional factor `f.medincome` is created to discretize the data according
to the quartiles.

```{r}
df$f.medincome <- discretize_quartiles(df$medincome, "MedianInc")
```

### Variable 6 - popest2015

Another continuous ratio variable not normally distributed.

```{r}
summary(df$popest2015)

hist(df$popest2015, breaks = 30, freq = F)
curve(dnorm(x, mean(df$popest2015), sd(df$popest2015)), add = T)

shapiro.test(df$popest2015)
```

An additional factor `f.popest2015` is created to discretize the data according
to the quartiles.

```{r}
df$f.popest2015 <- discretize_quartiles(df$popest2015, "MidPop")
```

### Variable 7 - povertypercent

Another continuous ratio variable not normally distributed. 

```{r}
summary(df$povertypercent)

hist(df$povertypercent, breaks = 30, freq = F)
curve(dnorm(x, mean(df$povertypercent), sd(df$povertypercent)), add = T)

shapiro.test(df$povertypercent)
```

An additional factor `f.povertypercent` is created to discretize the data
according to the quartiles.

```{r}
df$f.povertypercent <- discretize_quartiles(df$povertypercent, "Pov%")
```

### Variable 8 - studypercap

Another continuous ratio variable. This variable has the peculiarity of having a 
lot of 0s (median is also 0 so more than half of the counties don't perform 
cancer related clinical trials). It is not normally distributed.

```{r}
summary(df$studypercap)

hist(df$studypercap, breaks = 30, freq = F)
curve(dnorm(x, mean(df$studypercap), sd(df$studypercap)), add = T)

shapiro.test(df$studypercap)
```

An additional factor `f.studypercap` is created to discretize the data, this 
time groupping the data in only 3 groups: 0, and the two median splits of the
non-zero values.

```{r}
non_zero_median <- median(df$studypercap[df$studypercap > 0])

df$f.studypercap <- cut(df$studypercap, breaks = c(-Inf, 0, non_zero_median, Inf),
    include.lowest = T,
    labels=c("NoTrials", "MidTrials", "HighTrials")
  )
table(df$f.studypercap)
```

### Variable 9 - binnedinc

After having converted it from a string representation of the bin into a numeric
variable, analyzing its normality with Shapiro Test, we can safely say it's not
a normally-distributed variable.

```{r}
summary(df$binnedinc)

hist(df$binnedinc, breaks = 30, freq = F)
curve(dnorm(x, mean(df$binnedinc), sd(df$binnedinc)), add = T)

shapiro.test(df$binnedinc)
```

No further discretisation is needed for this variable, as it is already
categorised.

### Variable 10 - medianage

This is a continuous interval variable. By using a histogram we see that there 
are some data points that make no sense (median ages over 100), so the data is 
erroneous. Since we have data for male median age and female median age will 
clean the data by replacing the ouliers by the mean of male and female age. 
After cleaning the data the variable has no missing data, is not normal by means 
of the shapiro test and has 50 outliers (5 of them severe) in both ends of the 
spectrum. We create a factor variable "f.medianage" according to the quartiles. 

```{r}
summary(df$medianage)

hist(df$medianage, breaks = 30, freq = F)
curve(dnorm(x, mean(df$medianage), sd(df$medianage)), add = T)

df$medianage[df$medianage>100] <- (df$medianagemale[df$medianage > 100] + df$medianagefemale[df$medianage > 100]) / 2

summary(df$medianage)

hist(df$medianage, breaks = 30, freq = F)
curve(dnorm(x, mean(df$medianage), sd(df$medianage)), add = T)

shapiro.test(df$medianage)

sum(is.na(df$medianage))

Boxplot(df$medianage)

length(Boxplot(df$medianage, id = list(n=Inf)))

sevout_medianage = (quantile(df$medianage,0.25)+(3*((quantile(df$medianage,0.75)-quantile(df$medianage,0.25)))))
length(which(df$medianage > sevout_medianage))

df$f.medianage <- ifelse(df$medianage <= 37.85, 1, ifelse(df$medianage > 37.85 & df$medianage <= 40.90, 2, ifelse(df$medianage > 40.90 & df$medianage <= 43.90, 3, ifelse(df$medianage > 43.90, 4,0))))
df$f.medianage <- factor(df$f.medianage, labels=c("LowAge","LowMidAge","HighMidAge","HighAge"), order = T, levels=c(1,2,3,4))
table(df$f.medianage)
```


### Variable 11 - medianagemale

Very similar to the previous variable, this is a continuous interval variable, 
but with no apparent erroneous input. The variable has no missing data, is not 
normal by means of the shapiro test and has 46 outliers (6 of them severe) in 
both ends of the spectrum. We create a factor variable "f.medianagemale" 
according to the quartiles. The summary shows that male median age is slightly 
lower than median age (and thus lower than female median age). 

```{r}
summary(df$medianagemale)

hist(df$medianagemale, breaks = 30, freq = F)
curve(dnorm(x, mean(df$medianagemale), sd(df$medianagemale)), add = T)

shapiro.test(df$medianagemale)

sum(is.na(df$medianagemale))

Boxplot(df$medianagemale)

length(Boxplot(df$medianagemale, id = list(n=Inf)))

sevout_medianagemale = (quantile(df$medianagemale,0.25)+(3*((quantile(df$medianagemale,0.75)-quantile(df$medianagemale,0.25)))))
length(which(df$medianagemale > sevout_medianagemale))

df$f.medianagemale <- ifelse(df$medianagemale <= 36.40, 1, ifelse(df$medianagemale > 36.40 & df$medianagemale <= 39.50, 2, ifelse(df$medianagemale > 39.50 & df$medianagemale <= 42.60, 3, ifelse(df$medianagemale > 42.60, 4,0))))
df$f.medianagemale <- factor(df$f.medianagemale, labels=c("LowAgeMale","LowMidAgeMale","HighMidAgeMale","HighAgeMale"), order = T, levels=c(1,2,3,4))
table(df$f.medianagemale)
```


### Variable 12 - medianagefemale

We repeat the analysis for female median age. The variable has no apparent 
erroneous input, no missing data, is not normal by means of the shapiro test and 
has 55 outliers (1 of them severe) in both ends of the spectrum. We create a 
factor variable "f.medianagefemale" according to the quartiles. 

```{r}
summary(df$medianagefemale)

hist(df$medianagefemale, breaks = 30, freq = F)
curve(dnorm(x, mean(df$medianagefemale), sd(df$medianagefemale)), add = T)

shapiro.test(df$medianagefemale)

sum(is.na(df$medianagefemale))

Boxplot(df$medianagefemale)

length(Boxplot(df$medianagefemale, id = list(n=Inf)))

sevout_medianagefemale = (quantile(df$medianagefemale,0.25)+(3*((quantile(df$medianagefemale,0.75)-quantile(df$medianagefemale,0.25)))))
length(which(df$medianagefemale > sevout_medianagefemale))

df$f.medianagefemale <- ifelse(df$medianagefemale <= 39.20, 1, ifelse(df$medianagefemale > 39.20 & df$medianagefemale <= 42.40, 2, ifelse(df$medianagefemale > 42.40 & df$medianagefemale <= 45.30, 3, ifelse(df$medianagefemale > 45.30, 4,0))))
df$f.medianagefemale <- factor(df$f.medianagefemale, labels=c("LowAgeFemale","LowMidAgeFemale","HighMidAgeFemale","HighAgeFemale"), order = T, levels=c(1,2,3,4))
table(df$f.medianagefemale)

summary(df$geography)
```


### Variable 13 - geography

This is a string variable that is unique for each row of data. Since it is 
unique we could delete it, but it has info on not only the unique county of each 
observation, but also on its state. We will take this information and create a 
new variable named State that could be beneficial to our analysis. The new 
variable is a Nominal variable without missing values. However it has a lot of 
levels (50) with a few sparsly populated so it's not feasible to convert it to 
factor. 

```{r}
sample(df$geography, 10)

# Use regex to get the state (everything after the comma and white space):
df$state <- sub(".*,\\s*", "", df$geography)

summary(df$state)

table(df$state)

unique(df$state)

sum(is.na(df$state))

```


### Variable 13 - percentmarried

Another continuous ratio variable not normally distributed with 0 missing 
values, 34 outliers (none of them severe), all on the lower end. We create an 
additional ordinal factor “f.percentmarried”. 

```{r}
summary(df$percentmarried)

hist(df$percentmarried, breaks = 30, freq = F)
curve(dnorm(x, mean(df$percentmarried), sd(df$percentmarried)), add = T)

shapiro.test(df$percentmarried)

sum(is.na(df$percentmarried))

Boxplot(df$percentmarried)

length(Boxplot(df$percentmarried, id = list(n=Inf)))

sevout_percentmarried = (quantile(df$percentmarried,0.25)+(3*((quantile(df$percentmarried,0.75)-quantile(df$percentmarried,0.25)))))
length(which(df$percentmarried > sevout_percentmarried))

df$f.percentmarried <- ifelse(df$percentmarried <= 47.8, 1, ifelse(df$percentmarried > 47.8 & df$percentmarried <= 52.5, 2, ifelse(df$percentmarried > 52.5 & df$percentmarried <= 56.4, 3, ifelse(df$percentmarried > 56.4, 4,0))))
df$f.percentmarried <- factor(df$f.percentmarried, labels=c("LowMarriage%","LowMidMarriage%","HighMidMarriage%","HighMarriage%"), order = T, levels=c(1,2,3,4))
table(df$f.percentmarried)
```

### Variable 14 - pctnohs18_24

Another continuous ratio variable not normally distributed with 0 missing 
values, 34 outliers (none of them severe), all on the higher end. We create an 
additional ordinal factor “f.pctnohs18_24”. 

```{r}
summary(df$pctnohs18_24)

hist(df$pctnohs18_24, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pctnohs18_24), sd(df$pctnohs18_24)), add = T)

shapiro.test(df$pctnohs18_24)

sum(is.na(df$pctnohs18_24))

Boxplot(df$pctnohs18_24)

length(Boxplot(df$pctnohs18_24, id = list(n=Inf)))

sevout_pctnohs18_24 = (quantile(df$pctnohs18_24,0.25)+(3*((quantile(df$pctnohs18_24,0.75)-quantile(df$pctnohs18_24,0.25)))))
length(which(df$pctnohs18_24 > sevout_pctnohs18_24))

df$f.pctnohs18_24 <- ifelse(df$pctnohs18_24 <= 12.90, 1, ifelse(df$pctnohs18_24 > 12.90 & df$pctnohs18_24 <= 17.20, 2, ifelse(df$pctnohs18_24 > 17.20 & df$pctnohs18_24 <= 22.70, 3, ifelse(df$pctnohs18_24 > 22.70, 4,0))))
df$f.pctnohs18_24 <- factor(df$f.pctnohs18_24, labels=c("LowNoHighsc%","LowMidNoHighsc%","HighMidNoHighsc%","HighNoHighsc%"), order = T, levels=c(1,2,3,4))
table(df$f.pctnohs18_24)
```


### Variable 15 - pcths18_24

Another continuous ratio variable (related to the previous one) not normally 
distributed with 0 missing values, 33 outliers (9 of them severe) on both ends. 
There is one really severe outlier with 0 percent of High School Graduates, 
Greeley County, Kansas. It also has only 4.8% non High School Graduates (really 
low) and NA college graduates with a population of 1330. It seems like the 
values are probably false. For now we will leave it as such and later we will 
see how to deal with it. We create an additional ordinal factor “f.pcths18_24”. 

```{r}
summary(df$pcths18_24)

hist(df$pcths18_24, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pcths18_24), sd(df$pcths18_24)), add = T)

shapiro.test(df$pcths18_24)

sum(is.na(df$pcths18_24))

Boxplot(df$pcths18_24)

length(Boxplot(df$pcths18_24, id = list(n=Inf)))

sevout_pcths18_24 = (quantile(df$pcths18_24,0.25)+(3*((quantile(df$pcths18_24,0.75)-quantile(df$pcths18_24,0.25)))))
length(which(df$pcths18_24 > sevout_pcths18_24))

df[786,]

df$f.pcths18_24 <- ifelse(df$pcths18_24 <= 29.2, 1, ifelse(df$pcths18_24 > 29.2 & df$pcths18_24 <= 34.7, 2, ifelse(df$pcths18_24 > 34.7 & df$pcths18_24 <= 40.5, 3, ifelse(df$pcths18_24 > 40.5, 4,0))))
df$f.pcths18_24 <- factor(df$f.pcths18_24, labels=c("LowHighsc%","LowMidHighsc%","HighMidHighsc%","HighHighsc%"), order = T, levels=c(1,2,3,4))
table(df$f.pcths18_24)
```

### Variable 16 - pctsomecol18_24

Another continuous ratio variable (related to the 2 previous ones). It has 1376 
missing values which is more than 75% of our sample. This is too much and we 
will take the decision to take this variable out of the study because of with 
such a high proportion of missing data, it will not provide meaningful 
information. 

```{r}
summary(df$pctsomecol18_24)

hist(df$pctsomecol18_24, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pctsomecol18_24), sd(df$pctsomecol18_24)), add = T)

sum(is.na(df$pctsomecol18_24))

1376/1831*100


#Removing the column
df <- subset(df, select = -pctsomecol18_24)
```


### Variable 17 - pcths25_over

Another continuous ratio variable not normally distributed with 0 missing 
values, 18 outliers (none of them severe), all on the lower end. We create an 
additional ordinal factor “f.pcths25_over”. 

```{r}
summary(df$pcths25_over)

hist(df$pcths25_over, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pcths25_over), sd(df$pcths25_over)), add = T)

shapiro.test(df$pcths25_over)

sum(is.na(df$pcths25_over))

Boxplot(df$pcths25_over)

length(Boxplot(df$pcths25_over, id = list(n=Inf)))

sevout_pcths25_over = (quantile(df$pcths25_over,0.25)+(3*((quantile(df$pcths25_over,0.75)-quantile(df$pcths25_over,0.25)))))
length(which(df$pcths25_over > sevout_pcths25_over))

df$f.pcths25_over <- ifelse(df$pcths25_over <= 30.35, 1, ifelse(df$pcths25_over > 30.35 & df$pcths25_over <= 35.30, 2, ifelse(df$pcths25_over > 35.30 & df$pcths25_over <= 39.65, 3, ifelse(df$pcths25_over > 39.65, 4,0))))
df$f.pcths25_over <- factor(df$f.pcths25_over, labels=c("Low25Highsc%","LowMid25Highsc%","HighMid25Highsc%","High25Highsc%"), order = T, levels=c(1,2,3,4))
table(df$f.pcths25_over)
```


### Variable 18 - pctbachdeg25_over

Another continuous ratio variable (related to the previous one) not normally 
distributed with 0 missing values, 59 outliers (27 of them severe) all on the 
higher end. We create an additional ordinal factor “f.pctbachdeg25_over”. 

```{r}
summary(df$pctbachdeg25_over)

hist(df$pctbachdeg25_over, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pctbachdeg25_over), sd(df$pctbachdeg25_over)), add = T)

shapiro.test(df$pctbachdeg25_over)

sum(is.na(df$pctbachdeg25_over))

Boxplot(df$pctbachdeg25_over)

length(Boxplot(df$pctbachdeg25_over, id = list(n=Inf)))

sevout_pctbachdeg25_over = (quantile(df$pctbachdeg25_over,0.25)+(3*((quantile(df$pctbachdeg25_over,0.75)-quantile(df$pctbachdeg25_over,0.25)))))
length(which(df$pctbachdeg25_over > sevout_pctbachdeg25_over))

df$f.pctbachdeg25_over <- ifelse(df$pctbachdeg25_over <= 9.3, 1, ifelse(df$pctbachdeg25_over > 9.3 & df$pctbachdeg25_over <= 12.3, 2, ifelse(df$pctbachdeg25_over > 12.3 & df$pctbachdeg25_over <= 16.0, 3, ifelse(df$pctbachdeg25_over > 16.0, 4,0))))
df$f.pctbachdeg25_over <- factor(df$f.pctbachdeg25_over, labels=c("LowBach%","LowMidBach%","HighMidBach%","HighBach%"), order = T, levels=c(1,2,3,4))
table(df$f.pctbachdeg25_over)
```



### Variable 19 - pctemployed16_over

Another continuous ratio variable not normally distributed with 82 missing 
values (we will see how to input them later), 11 outliers (none of them severe), 
all but one on the lower end. We create an additional ordinal factor 
“f.pctemployed16_over”. 

```{r}
summary(df$pctemployed16_over)

hist(df$pctemployed16_over, breaks = 30, freq = F)

shapiro.test(df$pctemployed16_over)

sum(is.na(df$pctemployed16_over))

Boxplot(df$pctemployed16_over)

length(Boxplot(df$pctemployed16_over, id = list(n=Inf)))

sevout_pctemployed16_over = (48.60+(3*(60.30-48.60)))
length(which(df$pctemployed16_over > sevout_pctemployed16_over))

df$f.pctemployed16_over <- ifelse(df$pctemployed16_over <= 48.60, 1, ifelse(df$pctemployed16_over > 48.60 & df$pctemployed16_over <= 54.50, 2, ifelse(df$pctemployed16_over > 54.50 & df$pctemployed16_over <= 60.30, 3, ifelse(df$pctemployed16_over > 60.30, 4,0))))
df$f.pctemployed16_over <- factor(df$f.pctemployed16_over, labels=c("LowEmploy%","LowMidEmploy%","HighMidEmploy%","HighEmploy%"), order = T, levels=c(1,2,3,4))
table(df$f.pctemployed16_over)
```


### Variable 20 - pctunemployed16_over

One would assume that this variable is 100 minus the previous variable, but 
looking at some observations this is proven false. It is a continuous ratio 
variable not normally distributed with 0 missing values, 42 outliers (18 of them 
severe), all on the higher end. We create an additional ordinal factor 
“f.pctunemployed16_over”. 

```{r}
summary(df$pctunemployed16_over)

hist(df$pctunemployed16_over, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pctunemployed16_over), sd(df$pctunemployed16_over)), add = T)

shapiro.test(df$pctunemployed16_over)

sum(is.na(df$pctunemployed16_over))

Boxplot(df$pctunemployed16_over)

length(Boxplot(df$pctunemployed16_over, id = list(n=Inf)))

sevout_pctunemployed16_over = (quantile(df$pctunemployed16_over,0.25)+(3*((quantile(df$pctunemployed16_over,0.75)-quantile(df$pctunemployed16_over,0.25)))))
length(which(df$pctunemployed16_over > sevout_pctunemployed16_over))

df$f.pctunemployed16_over <- ifelse(df$pctunemployed16_over <= 5.5, 1, ifelse(df$pctunemployed16_over > 5.5 & df$pctunemployed16_over <= 7.5, 2, ifelse(df$pctunemployed16_over > 7.5 & df$pctunemployed16_over <= 9.75, 3, ifelse(df$pctunemployed16_over > 9.75, 4,0))))
df$f.pcuntemployed16_over <- factor(df$f.pctunemployed16_over, labels=c("LowUnEmploy%","LowMidUnEmploy%","HighMidUnEmploy%","HighUnEmploy%"), order = T, levels=c(1,2,3,4))
table(df$f.pctunemployed16_over)
```


### Variable 21 - pctprivatecoverage

Another continuous ratio variable not normally distributed with 0 missing 
values, 17 outliers (none of them severe) all on the lower end. We create an 
additional ordinal factor “f.pctprivatecoverage”. 

```{r}
summary(df$pctprivatecoverage)

hist(df$pctprivatecoverage, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pctprivatecoverage), sd(df$pctprivatecoverage)), add = T)

shapiro.test(df$pctprivatecoverage)

sum(is.na(df$pctprivatecoverage))

Boxplot(df$pctprivatecoverage)

length(Boxplot(df$pctprivatecoverage, id = list(n=Inf)))

sevout_pctprivatecoverage = (quantile(df$pctprivatecoverage,0.25)+(3*((quantile(df$pctprivatecoverage,0.75)-quantile(df$pctprivatecoverage,0.25)))))
length(which(df$pctprivatecoverage > sevout_pctprivatecoverage))

df$f.pctprivatecoverage <- ifelse(df$pctprivatecoverage <= 57.50, 1, ifelse(df$pctprivatecoverage > 57.50 & df$pctprivatecoverage <= 65.20, 2, ifelse(df$pctprivatecoverage > 65.20 & df$pctprivatecoverage <= 72.10, 3, ifelse(df$pctprivatecoverage > 72.10, 4,0))))
df$f.pctprivatecoverage <- factor(df$f.pctprivatecoverage, labels=c("LowPrivate%","LowMidPrivate%","HighMidPrivate%","HighPrivate%"), order = T, levels=c(1,2,3,4))
table(df$f.pctprivatecoverage)
```


### Variable 22 - pctprivatecoveragealone

This is a continuous ratio variable very closely related with the previous 
variable. It also has 356 missing values, which amounts to almost 20% of the 
observations. Since the number of missing values is high and it doesn't add much 
to our data (it has a correlation of 0.93 with the previous variable) we will 
delete it. 

```{r}
summary(df$pctprivatecoveragealone)

sum(is.na(df$pctprivatecoveragealone))

356/1831*100

cor.test(df$pctprivatecoverage, df$pctprivatecoveragealone)

df <- subset(df, select = -pctprivatecoveragealone)
```


### Variable 22 - pctempprivcoverage

Another continuous ratio variable normally distributed (if we pick a 99% 
significance level for the shapiro test) with 0 missing values, 7 outliers (none 
of them severe) all on the higher end but one. We create an additional ordinal 
factor “f.pctempprivcoverage”. 

```{r}
summary(df$pctempprivcoverage)

hist(df$pctempprivcoverage, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pctempprivcoverage), sd(df$pctempprivcoverage)), add = T)

shapiro.test(df$pctempprivcoverage)

sum(is.na(df$pctempprivcoverage))

Boxplot(df$pctempprivcoverage)

length(Boxplot(df$pctempprivcoverage, id = list(n=Inf)))

sevout_pctempprivcoverage = (quantile(df$pctempprivcoverage,0.25)+(3*((quantile(df$pctempprivcoverage,0.75)-quantile(df$pctempprivcoverage,0.25)))))
length(which(df$pctempprivcoverage > sevout_pctempprivcoverage))

df$f.pctempprivcoverage <- ifelse(df$pctempprivcoverage <= 34.60, 1, ifelse(df$pctempprivcoverage > 34.60 & df$pctempprivcoverage <= 41.10, 2, ifelse(df$pctempprivcoverage > 41.10 & df$pctempprivcoverage <= 47.70, 3, ifelse(df$pctempprivcoverage > 47.70, 4,0))))
df$f.pctempprivcoverage <- factor(df$f.pctempprivcoverage, labels=c("LowEmployeeHealth%","LowMidEmployeeHealth%","HighMidEmployeeHealth%","HighEmployeeHealth%"), order = T, levels=c(1,2,3,4))
table(df$f.pctempprivcoverage)
```


### Variable 23 - pctpubliccoverage

Another continuous ratio variable normally distributed with 0 missing values, 13 
outliers (1 of them severe) on both ends of the spectrum. We create an 
additional ordinal factor “f.pctpubliccoverage”. 

```{r}
summary(df$pctpubliccoverage)

hist(df$pctpubliccoverage, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pctpubliccoverage), sd(df$pctpubliccoverage)), add = T)

shapiro.test(df$pctpubliccoverage)

sum(is.na(df$pctpubliccoverage))

Boxplot(df$pctpubliccoverage)

length(Boxplot(df$pctpubliccoverage, id = list(n=Inf)))

sevout_pctpubliccoverage = (quantile(df$pctpubliccoverage,0.25)+(3*((quantile(df$pctpubliccoverage,0.75)-quantile(df$pctpubliccoverage,0.25)))))
length(which(df$pctpubliccoverage > sevout_pctpubliccoverage))

df$f.pctpubliccoverage <- ifelse(df$pctpubliccoverage <= 30.90, 1, ifelse(df$pctpubliccoverage > 30.90 & df$pctpubliccoverage <= 36.30, 2, ifelse(df$pctpubliccoverage > 36.30 & df$pctpubliccoverage <= 41.40, 3, ifelse(df$pctpubliccoverage > 41.40, 4,0))))
df$f.pctpubliccoverage <- factor(df$f.pctpubliccoverage, labels=c("LowGovHealth%","LowMidGovHealth%","HighMidGovHealth%","HighGovHealth%"), order = T, levels=c(1,2,3,4))
table(df$f.pctpubliccoverage)
```


### Variable 24 - pctpubliccoveragealone

Another continuous ratio variable related to the previous variable (this time 
with no NAs and not as closely correlated as variables 21 and 22, cor=0.87, so 
we will keep de variable for now) not normally distributed with 0 missing 
values, 21 outliers (7 of them severe) on the higher end (except one). We create 
an additional ordinal factor “f.pctpubliccoveragealone”. 

```{r}
summary(df$pctpubliccoveragealone)

cor.test(df$pctpubliccoverage, df$pctpubliccoveragealone)

hist(df$pctpubliccoveragealone, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pctpubliccoveragealone), sd(df$pctpubliccoveragealone)), add = T)

shapiro.test(df$pctpubliccoveragealone)

sum(is.na(df$pctpubliccoveragealone))

Boxplot(df$pctpubliccoveragealone)

length(Boxplot(df$pctpubliccoveragealone, id = list(n=Inf)))

sevout_pctpubliccoveragealone = (quantile(df$pctpubliccoveragealone,0.25)+(3*((quantile(df$pctpubliccoveragealone,0.75)-quantile(df$pctpubliccoveragealone,0.25)))))
length(which(df$pctpubliccoveragealone > sevout_pctpubliccoveragealone))

df$f.pctpubliccoveragealone <- ifelse(df$pctpubliccoveragealone <= 14.90, 1, ifelse(df$pctpubliccoveragealone > 14.90 & df$pctpubliccoveragealone <= 18.70, 2, ifelse(df$pctpubliccoveragealone > 18.70 & df$pctpubliccoveragealone <= 23.00, 3, ifelse(df$pctpubliccoveragealone > 23.00, 4,0))))
df$f.pctpubliccoveragealone <- factor(df$f.pctpubliccoveragealone, labels=c("LowGovHealthAlone%","LowMidGovHealthAlone%","HighMidGovHealthAlone%","HighGovHealthAlone%"), order = T, levels=c(1,2,3,4))
table(df$f.pctpubliccoveragealone)
```


### Variable 25 - pctwhite

Another continuous ratio variable clearly not normally distributed with 0 
missing values, 97 outliers (none of them severe) all on the low end of the 
spectrum. We create an additional ordinal factor “f.pctwhite”. 

```{r}
summary(df$pctwhite)

hist(df$pctwhite, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pctwhite), sd(df$pctwhite)), add = T)

shapiro.test(df$pctwhite)

sum(is.na(df$pctwhite))

Boxplot(df$pctwhite)

length(Boxplot(df$pctwhite, id = list(n=Inf)))

sevout_pctwhite = (quantile(df$pctwhite,0.25)+(3*((quantile(df$pctwhite,0.75)-quantile(df$pctwhite,0.25)))))
length(which(df$pctwhite > sevout_pctwhite))

df$f.pctwhite <- ifelse(df$pctwhite <= 77.31, 1, ifelse(df$pctwhite > 77.31 & df$pctwhite <= 89.90, 2, ifelse(df$pctwhite > 89.90 & df$pctwhite <= 95.57, 3, ifelse(df$pctwhite > 95.57, 4,0))))
df$f.pctwhite <- factor(df$f.pctwhite, labels=c("LowWhite%","LowMidWhite%","HighMidWhite%","HighWhite%"), order = T, levels=c(1,2,3,4))
table(df$f.pctwhite)
```


### Variable 26 - pctblack

Really similar to the previous variable, with a correlation of 0.84. It is 
another continuous ratio variable clearly not normally distributed with 0 
missing values, 224 outliers (168 of them severe) all on the high end of the 
spectrum. We create an additional ordinal factor “f.pctblack”. 

```{r}
summary(df$pctblack)

cor.test(df$pctwhite, df$pctblack)

hist(df$pctblack, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pctblack), sd(df$pctblack)), add = T)

shapiro.test(df$pctblack)

sum(is.na(df$pctblack))

Boxplot(df$pctblack)

length(Boxplot(df$pctblack, id = list(n=Inf)))

sevout_pctblack = (quantile(df$pctblack,0.25)+(3*((quantile(df$pctblack,0.75)-quantile(df$pctblack,0.25)))))
length(which(df$pctblack > sevout_pctblack))

df$f.pctblack <- ifelse(df$pctblack <= 0.648, 1, ifelse(df$pctblack > 0.648 & df$pctblack <= 2.323, 2, ifelse(df$pctblack > 2.323 & df$pctblack <= 10.867, 3, ifelse(df$pctblack > 10.867, 4,0))))
df$f.pctblack <- factor(df$f.pctblack, labels=c("LowBlack%","LowMidBlack%","HighMidBlack%","HighBlack%"), order = T, levels=c(1,2,3,4))
table(df$f.pctblack)
```


### Variable 27 - pctasian

Also related to the previous 2 variables. It is a continuous ratio variable 
clearly not normally distributed with 0 missing values, 198 outliers (156 of 
them severe, and looking at the boxplot some of them really far, probably asian 
ghetto counties) all on the high end of the spectrum. We create an additional 
ordinal factor “f.pctasian”. 

```{r}
summary(df$pctasian)

hist(df$pctasian, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pctasian), sd(df$pctasian)), add = T)

shapiro.test(df$pctasian)

sum(is.na(df$pctasian))

Boxplot(df$pctasian)

length(Boxplot(df$pctasian, id = list(n=Inf)))

sevout_pctasian = (quantile(df$pctasian,0.25)+(3*((quantile(df$pctasian,0.75)-quantile(df$pctasian,0.25)))))
length(which(df$pctasian > sevout_pctasian))

df$f.pctasian <- ifelse(df$pctasian <= 0.2582, 1, ifelse(df$pctasian > 0.2582 & df$pctasian <= 0.5495, 2, ifelse(df$pctasian > 0.5495 & df$pctasian <= 1.2515, 3, ifelse(df$pctasian > 1.2515, 4,0))))
df$f.pctasian <- factor(df$f.pctasian, labels=c("LowAsian%","LowMidAsian%","HighMidAsian%","HighAsian%"), order = T, levels=c(1,2,3,4))
table(df$f.pctasian)
```


### Variable 28 - pctotherrace

This variable should be 100 minus the sum of the three previous variables but 
looking at a sample of observations it is clearly not, and also if we check for 
multicollinearity using VIF, since the values are lower than 5 we can use the 
rule of thumb to say that there is not a severe multicollinearity so we will 
keep the variable for now (if it was always equal to 100 we would erase it since 
it wouldn't add any new info). 
The variable is a continuous ratio variable clearly not normally distributed 
with 0 missing values, 181 outliers (148 of them severe, and looking at the 
boxplot some of them really far, probably asian ghetto counties) all on the high 
end of the spectrum. We create an additional ordinal factor “f.pctotherrace”. 

```{r}
summary(df$pctotherrace)

model <- lm(pctotherrace ~ pctwhite + pctblack + pctasian, data=df)
vif(model)

summary(df$pctotherrace)

hist(df$pctotherrace, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pctotherrace), sd(df$pctotherrace)), add = T)

shapiro.test(df$pctotherrace)

sum(is.na(df$pctotherrace))

Boxplot(df$pctotherrace)

length(Boxplot(df$pctotherrace, id = list(n=Inf)))

sevout_pctotherrace = (quantile(df$pctotherrace,0.25)+(3*((quantile(df$pctotherrace,0.75)-quantile(df$pctotherrace,0.25)))))
length(which(df$pctotherrace > sevout_pctotherrace))

df$f.pctotherrace <- ifelse(df$pctotherrace <= 0.2867, 1, ifelse(df$pctotherrace > 0.2867 & df$pctotherrace <= 0.7826, 2, ifelse(df$pctotherrace > 0.7826 & df$pctotherrace <= 2.1066, 3, ifelse(df$pctotherrace > 2.1066, 4,0))))
df$f.pctotherrace <- factor(df$f.pctotherrace, labels=c("LowOtherRace%","LowMidOtherRace%","HighMidOtherRace%","HighOtherRace%"), order = T, levels=c(1,2,3,4))
table(df$f.pctotherrace)
```


### Variable 29 - pctmarriedhouseholds

Another continuous ratio variable not normally distributed with 0 missing 
values, 57 outliers (2 of them severe) on both ends of the spectrum. We create 
an additional ordinal factor “f.pctmarriedhouseholds”. 

```{r}
summary(df$pctmarriedhouseholds)

hist(df$pctmarriedhouseholds, breaks = 30, freq = F)
curve(dnorm(x, mean(df$pctmarriedhouseholds), sd(df$pctmarriedhouseholds)), add = T)

shapiro.test(df$pctmarriedhouseholds)

sum(is.na(df$pctmarriedhouseholds))

Boxplot(df$pctmarriedhouseholds)

length(Boxplot(df$pctmarriedhouseholds, id = list(n=Inf)))

sevout_pctmarriedhouseholds = (quantile(df$pctmarriedhouseholds,0.25)+(3*((quantile(df$pctmarriedhouseholds,0.75)-quantile(df$pctmarriedhouseholds,0.25)))))
length(which(df$pctmarriedhouseholds > sevout_pctmarriedhouseholds))

df$f.pctmarriedhouseholds <- ifelse(df$pctmarriedhouseholds <= 47.85, 1, ifelse(df$pctmarriedhouseholds > 47.85 & df$pctmarriedhouseholds <= 51.73, 2, ifelse(df$pctmarriedhouseholds > 51.73 & df$pctmarriedhouseholds <= 55.48, 3, ifelse(df$pctmarriedhouseholds > 55.48, 4,0))))
df$f.pctmarriedhouseholds <- factor(df$f.pctmarriedhouseholds, labels=c("LowMarried%","LowMidMarried%","HighMidMarried%","HighMarried%"), order = T, levels=c(1,2,3,4))
table(df$f.pctmarriedhouseholds)
```


### Variable 30 - birthrate

The last variable is yet another continuous ratio variable not normally 
distributed with 0 missing values, 104 outliers (52 of them severe) on both ends 
of the spectrum. We create an additional ordinal factor “f.birthrate”. 

```{r}
summary(df$birthrate)

hist(df$birthrate, breaks = 30, freq = F)
curve(dnorm(x, mean(df$birthrate), sd(df$birthrate)), add = T)

shapiro.test(df$birthrate)

sum(is.na(df$birthrate))

Boxplot(df$birthrate)

length(Boxplot(df$birthrate, id = list(n=Inf)))

sevout_birthrate = (quantile(df$birthrate,0.25)+(3*((quantile(df$birthrate,0.75)-quantile(df$birthrate,0.25)))))
length(which(df$birthrate > sevout_birthrate))

df$f.birthrate <- ifelse(df$birthrate <= 4.528, 1, ifelse(df$birthrate > 4.528 & df$birthrate <= 5.355, 2, ifelse(df$birthrate > 5.355 & df$birthrate <= 6.414, 3, ifelse(df$birthrate > 6.414, 4,0))))
df$f.birthrate <- factor(df$f.birthrate, labels=c("LowBirth%","LowMidBirth%","HighMidBirth%","HighBirth%"), order = T, levels=c(1,2,3,4))
table(df$f.birthrate)
```



## Autocorrelation

```{r}
acf(df$target_deathrate)
```


## Profiling
```{r}
 res.con = condes(df[, c(1:7, 9:12, 14:31)],3)

correlation_matrix <- res.con$quanti
correlation_values <- correlation_matrix[, "correlation"]
variable_names <- rownames(correlation_matrix)

correlation_df <- data.frame(
  variable = variable_names,
  correlation = correlation_values
)

filtered_correlation_df <- correlation_df[abs(correlation_df$correlation) > 0.35, ]

sorted_correlation_df <- filtered_correlation_df[order(abs(filtered_correlation_df$correlation), decreasing = TRUE), ]

print(sorted_correlation_df[,"variable"])

```

## Check multicolinearity and low correlation with the target (remove or combine variables)

First,we must identify those variables that are very correlated and combine 
them. We do so because high correlations (close to 1 or -1) between two or more 
predictors indicate potential multicollinearity. Initially, we combine or 
eliminate those variables with a correlation above 0.8 or below -0.8. 

```{r}

cor_matrix <- cor(df[, c(1:7, 9:12, 14:31)])

corrplot(cor_matrix, method = "circle")

#Identify which varaibles present a correlation higher than 0.8 or lower than -0.8
high_corr_vars <- c()

for(i in 1:ncol(cor_matrix)){
  for(j in i:ncol(cor_matrix)){
    if(abs(cor_matrix[i,j]) > 0.9 & i != j){
      # Append the variable name to remove (either i or j)
      high_corr_vars <- c(high_corr_vars, colnames(cor_matrix)[j])
    }
  }
}

high_corr_vars <- unique(high_corr_vars)
print("Highly multicollinearity variables")
print(high_corr_vars)
print("Varialbes highly corralated with the target")
print(high_correlation)

#binnedinc highly corralated with medincome

#Medianage, medianagemale, medianagefemale can be treated as the same value as their correlation is very high. They are almost the same variable. Also all three of them very poorly correlated with the target

#Remove non-correlated variables with the target variable
cor_with_target <- cor_matrix[, "target_deathrate"]
vars_to_remove <- names(cor_with_target[abs(cor_with_target) < 0.1])

#Remove these variables from the dataset
df_cleaned <- df[, !(colnames(df) %in% vars_to_remove)]

#######
# pcths18_24 with pcths25_over
df_cleaned$pcths <- (df$pcths18_24 + df$pcths25_over) / 2

# pctbachdeg18_24 with pctbachdeg25_over
df_cleaned$pctbach <- (df$pctbachdeg18_24 + df$pctbachdeg25_over) / 2

df_cleaned <- df_cleaned[, !colnames(df_cleaned) %in% c("pcths18_24", "pcths25_over", "pctbachdeg18_24", "pctbachdeg25_over")]

#Combine Income-Related Variables
#df_cleaned$pctincome <- df_cleaned$binnedinc
#df_cleaned <- df_cleaned[, !colnames(df) %in% c("medincome")]

# Combine Race and Ethnicity Variables
#df_cleaned$racindex <- df_cleaned$pctblack + df$pctasian
#df_cleaned <- df_cleaned[, !colnames(df) %in% c("pctwhite")]

#Combine Public Coverage and Poverty Variables
#df_cleaned$social_welfare <- df_cleaned$pctpubliccoverage + df$povertypercent
#df_cleaned <- df_cleaned[, !colnames(df) %in% c("pctpubliccoverage", "povertypercent")]

#Combine or remove variables
#df_cleaned$medianage_combined <- (df_cleaned$medianage + df_cleaned$medianagemale + df_cleaned$medianagefemale) / 3

#df_cleaned <- df_cleaned[, !(colnames(df_cleaned) %in% c("medianage", "medianagemale"))]

new_cor_matrix <- cor(df_cleaned[, c(1:8, 10:21)])
new_cor_matrix[, "target_deathrate"]
corrplot(new_cor_matrix, method = "circle")

```

```{r}
#Most corralated +incidencerate  -medincome +povertypercent  +pcths25_over -pctemployed16_over pctunemployed16_over  pctprivatecoverage pctpubliccoverage pctpubliccoveragealone
model <- lm(target_deathrate ~ percentmarried * incidencerate  + pcths25_over +
              povertypercent + pctbachdeg25_over + pctemployed16_over + medincome * pcths18_24 + pctbachdeg18_24 * pcths25_over +
                   f.pctpubliccoveragealone, data = df_cleaned)

summary(model)

qqnorm(model$residuals)
qqline(model$residuals, col = "red")

# Breusch-Pagan test for heteroscedasticity
bptest(model)

lm_stepwise <- step(model, direction = "both", trace = 0)

vif(lm_stepwise)  

summary(lm_stepwise)
```



#Analyze behaviour of what seem to be the main predictors

```{r}
#Boxplot most relevant predictors
boxplot(target_deathrate ~ f.incidencerate, data = df_cleaned, main = "Death Rate vs. Incidence Rate")
boxplot(target_deathrate ~ f.medincome, data = df_cleaned, main = "Death Rate vs. Median Income")
boxplot(target_deathrate ~ f.percentmarried, data = df_cleaned, main = "Death Rate vs. Percent Married")
boxplot(target_deathrate ~ f.pcths18_24, data = df_cleaned, main = "Death Rate vs. Percent 18-24 HS")
boxplot(target_deathrate ~ f.povertypercent, data = df_cleaned, main = "Death Rate vs. Percent 18-24 HS")
boxplot(target_deathrate ~ f.pctpubliccoverage, data = df_cleaned, main = "Death Rate vs. Percent 18-24 HS")
boxplot(target_deathrate ~ f.pctpubliccoveragealone, data = df_cleaned, main = "Death Rate vs. Percent 18-24 HS")

#Pairwise tests
#The ANOVA result confirms that marital status percentage (f.percentmarried) has a significant effect on the target_deathrate, suggesting that different levels of marriage percentages correspond to different death rate levels.
pairwise.wilcox.test(df_cleaned$target_deathrate, df_cleaned$f.percentmarried)
oneway.test(target_deathrate ~ f.percentmarried, data = df_cleaned)

# Incidence Rate
#ANOVA results indicate that the f.incidencerate factor has a statistically significant impact on target_deathrate.
pairwise.wilcox.test(df_cleaned$target_deathrate, df_cleaned$f.incidencerate)
oneway.test(target_deathrate ~ f.incidencerate, data = df_cleaned)

# Median Income
# ANOVA tests indicate that f.medincome (median income) has a statistically significant impact on target_deathrate. Specifically, different levels of median income (such as LowMedianInc, LowMidMedianInc, etc.) are associated with different death rates, with higher income groups generally showing lower death rates.
pairwise.wilcox.test(df_cleaned$target_deathrate, df_cleaned$f.medincome)
oneway.test(target_deathrate ~ f.medincome, data = df_cleaned)

# Percent 18-24 HS
#ANOVA indicate a significant relationship between f.pcths18_24 (the percentage of people with high school education in the 18-24 age range) and target_deathrate. The results suggest that as the percentage of high school education increases, the death rates tend to decrease
pairwise.wilcox.test(df_cleaned$target_deathrate, df_cleaned$f.pcths18_24)
oneway.test(target_deathrate ~ f.pcths18_24, data = df_cleaned)

pairwise.wilcox.test(df_cleaned$target_deathrate, df_cleaned$f.pctpubliccoverage)
oneway.test(target_deathrate ~ f.pctpubliccoverage, data = df_cleaned)

pairwise.wilcox.test(df_cleaned$target_deathrate, df_cleaned$f.pctpubliccoverage)
oneway.test(target_deathrate ~ f.pctpubliccoverage, data = df_cleaned)

```

# Model Fitting

## No-interaction models

```{r}
# Linear model: target_deathrate ~ f.percentmarried --> 5.47%
lm_percentmarried <- lm(target_deathrate ~ f.percentmarried, data = df_cleaned)
summary(lm_percentmarried)

# Linear model: target_deathrate ~ f.incidencerate --> 15%
lm_incidencerate <- lm(target_deathrate ~ f.incidencerate, data = df_cleaned)
summary(lm_incidencerate)

# Linear model: target_deathrate ~ f.medincome --> 19%
lm_medincome <- lm(target_deathrate ~ f.medincome, data = df_cleaned)
summary(lm_medincome)

# Linear model: target_deathrate ~ f.pcths18_24 --> 5.1%
lm_pcths18_24 <- lm(target_deathrate ~ f.pcths18_24, data = df_cleaned)
summary(lm_pcths18_24)

# Linear model: target_deathrate ~ f.pctpubliccoverage --> 15%
lm_pctpubliccoverage <- lm(target_deathrate ~ f.pctpubliccoverage, data = df_cleaned) 
summary(lm_pctpubliccoverage)

# Linear model: target_deathrate ~ f.pctpubliccoveragealone --> 18.85%
lm_pctpubliccoveragealone <- lm(target_deathrate ~ f.pctpubliccoveragealone, data = df_cleaned) 
summary(lm_pctpubliccoveragealone)


```

## Interaction models 


```{r}
m1 <- lm(target_deathrate ~ f.medincome, data = df)
summary(m1)
par(mar = c(5, 5, 4, 2)) 
boxcox(target_deathrate ~ f.medincome , data=df) #No transformation of the target needed

m2 <- lm(target_deathrate ~ f.medincome + f.incidencerate, data= df)
boxcox(target_deathrate ~ f.medincome + f.incidencerate, data=df) #No transformation of the target needed
summary(m2)
t <- summary(m2)
vif(m2)
1/(1-t$r.squared)
  
anova(m1,m2)

#########################Fins aquí estic bastant segur que está be, no hi ha collinearity i el imporvement es considerable#########

m3 <- lm(target_deathrate ~ medincome + incidencerate + pctpubliccoveragealone, data= df)
summary(m3)

anova(m1,m3)

m4 <- lm(target_deathrate ~ medincome + incidencerate + pctbachdeg25_over + pctpubliccoveragealone, data= df)
summary(m4)

anova(m1,m4)

m5 <- lm(target_deathrate ~ medincome + incidencerate + pctbachdeg25_over + pctpubliccoveragealone + pctemployed16_over, data= df)
summary(m5)

anova(m4,m5)

m6 <- lm(target_deathrate ~ medincome + incidencerate + pctbachdeg25_over + pctpubliccoveragealone + pctemployed16_over + povertypercent  , data= df)
summary(m6)

anova(m5,m6)

m7 <- lm(target_deathrate ~ medincome + incidencerate + pctbachdeg25_over + pctpubliccoveragealone + pctemployed16_over + povertypercent 
         , data= df)
t <- summary(m7)
anova(m6,m7)
boxcox(target_deathrate ~ medincome + incidencerate + pctbachdeg25_over + pctpubliccoveragealone + pctemployed16_over + povertypercent, data=df)
vif(m7)

1/(1-t$r.squared)

```


```{r}
mm1 <- lm(target_deathrate ~ medincome + incidencerate + pctpubliccoverage + pctpubliccoveragealone + f.pcths18_24 + percentmarried, data= df_cleaned)

mm2 <- lm(target_deathrate ~ medincome + pctbachdeg25_over + pctpubliccoveragealone + povertypercent
         + incidencerate + pctemployed16_over + pctpubliccoverage + pctprivatecoverage + binnedinc + pctunemployed16_over + pctunemployed16_over
         , data = df)

vif(mm1)
```









# Model Validation